% -*- latex -*-

\section{Naming files and directories in the file system}

\subsection{Motivation}

The main difficulty with file naming lies in the fact that files or
even whole directories may move after make7 has already partially (but
not fully) processed them.  For example, this happens when the {\em
autoloader} (see Section~\REF{sec:autoload}) has been invoked and the
session (including make7's internal state) is then frozen (i.e., saved to
a file) via {\tt lib7.fork_to_disk}.

make7's configurable {\em path anchor} mechanism enables it to resume
such a session even when operating in a different environment, perhaps
on a different machine with different file systems mounted, or a
different location of the compiler installation.  Evaluation of path
anchors always takes place as late as possible, and make7 will re-evaluate
path anchors as this becomes necessary due to changes to their
configuration.

\subsection{Basic rules}
\label{sec:basicrules}

make7 uses its own ``standard'' syntax for pathnames which for the most
part happens to be the same as the one used by most Unix-like systems:
\begin{itemize}
\item Path name components are separated by ``{\bf /}''.
\item Special components ``{\bf .}'' and ``{\bf ..}'' denote {\em
current} and {\em previous} directory, respectively.
\item Paths beginning
with ``{\bf /}'' are considered {\em absolute}.
\item Other paths are {\em relative} unless they start with ``{\bf \$}''.
\end{itemize}
\noindent There is an important third form of standard paths: {\em
anchored} paths.  Anchored paths always start with ``{\bf \$}''.

Since this standard syntax does not cover system-specific aspects such
as volume names, it is also possible to revert to ``native'' syntax by
enclosing a path name in double-quotes.  Of course, description files
that use path names in native syntax are not portable across operating
systems.

\begin{description}
\item[Absolute pathnames] are resolved in the usual manner
specific to the operating system.  However, it is advisable to avoid
absolute pathnames because they are certain to ``break'' if the
corresponding file moves to a different location.
\item[Relative pathnames that occur in some make7 description file] whose
name is {\it path}{\tt /}{\it file}{\tt .make6} will be resolved relative
to {\it path}, i.e., relative to the directory that contains the
description file.
\item[Relative pathnames that have been entered interactively,]
usually as an argument to one of make7's interface functions, will be
resolved in the OS-specific manner, i.e., relative to the current
working directory.  However, notice that some of make7's operations (see
section~\REF{sec:autoload}---autoload) will be executed lazily and,
thus, can occur interleaved with arbitary other operations---including
changes of the working directory.  This is handled by make7 in such a way
that it appears as if all path derived from an interactive relative
path $p$ had been completely resolved at the time $p$ was entered. As
a result, two names specified using identical strings but at different
times when different working directories were in effect will be kept
apart and continue to refer to their respective original file system
locations.
\item[Anchored paths] consist of an anchor name (of non-zero length)
and a non-empty list of additional arcs.  The name is enclosed by
the path's leading {\bf \$} on the left and the path's first {\bf /}
on the right.  The list of arcs follows the first {\bf /}.  As with
all standard paths, the arcs themselves are also separated by {\bf /}.
An error is signalled if the anchor name is not known to make7.
If $a$ is a know anchor name currently bound to some directory name
$d$, then the standard path {\tt \$}$a${\tt /}$p$ (where $p$ is a list
of arcs) refers to $d${\tt /}$p$.  The frequently occuring case where
$a$ coincides with the first arc of $p$ can be abbreviated as {\tt
\$/}$p$.
\end{description}

\subsection{Anchor environments}
\label{sec:anchor:env}

Anchor names are resolved in the {\em anchor environment} that is in
effect at the time the anchor is read.

The basis for all anchor environments is the {\em root environment}.
Conceptually, the root environments is a fixed mapping that binds
every possible anchor to a mutable location.  The location can store a
native directory name or can be marked ``undefined''.  Most locations
initially start out undefined.  The contents of each location is
configurable (see Section~\REF{sec:anchor:config}).

At the time a make7 description file $a${\tt .make6} refers to another
library's or library component's description file $b${\tt .make6}, it can
augment the current anchor environment with new namings.  The new
namings are in effect while $b${\tt .make6} (including any description
files {\it it}\/ mentions!) is being processed.  If a new naming
binds an anchor name that was already bound in the current
environment\footnote{which is technically always the case given our
explanation of the root environment}, then the old naming is being
hidden.  The effect is scoping for anchor names.

Using make7's {\em tool parameter} mechanism (see
Section~\REF{sec:toolparam}), a new naming is specified as a pair of
anchor name and anchor value.  The value has the form of another path
name (standard or native). Example:

\begin{verbatim}
  a.make6 (bind:(anchor:lib value:$mysrc/a-lib)
        bind:(anchor:support value:$lib)
        bind:(anchor:utils value:/home/bob/src/ML/utils))
\end{verbatim}

As shown in this example, it is perfectly legal for the specification
of the value to involve the use of another anchor.  That anchor will
be resolved in the original anchor environment. Thus, a path anchored
at {\tt \$lib} in {\tt a.make6} will be resolved using the naming for
{\tt \$mystuff} that is currently in effect.  The point here is that a
re-configuration of the root environment that affects {\tt \$mystuff}
now also affects how {\tt \$lib} is resolved as it occurs within {\tt
a.make6}.

make7 processes the list of {\tt bind}-directives ``in parallel.'' This
means that the anchor {\tt \$support} will refer to the original
meaning of {\tt \$lib} and is {\em not} being bound to {\tt
\$mysrc/a-lib/asupport}.

The example also demonstrates that {\tt value}-paths can be single
anchors. In other words, the restriction that there has to be at least
one arc after the anchor does not apply here. This makes it possible
to ``rename'' anchors, or, to put it more precisely, for one anchor
name to be established as an ``alias'' for another anchor name.

\subsection{Anchor configuration}
\label{sec:anchor:config}

Anchor configuration is concerned with the values that are stored in
the root anchor environment.  At startup time, the root environment is
initialized by reading two configuration files: an
installation-specific one and a user-specific one.  After that, the
contents of root locations can be maintained using make7's interface
functions {\tt make7.anchor.anchor} and {\tt make7.anchor.reset} (see
Section~\REF{sec:api:anchors}).

Although there is a hard-wired default for the installation-specific
configuration file\footnote{which happens to be {\tt
/usr/lib/smlnj-pathconfig}}, this default is rarely used.
Instead, in a typical installation of the compiler the default will be a
file $r${\tt /lib/pathconfig} where $r$ is the {\it root} directory
into which the compiler had been installed.  (The installation procedure
establishes this new default by setting the environment variable {\tt
CM\_PATHCONFIG} at the time it produces the heap image for
the interactive system.)  The user can specify a new location at
startup time using the same environment variable {\tt CM\_PATHCONFIG}.

The default location of the user-specific configuration file is {\tt
.smlnj-pathconfig} in the user's home directory (which must be given
by the {\tt HOME} environment variable).  At startup time, this
default can be overridden by a fixed location which must be given as
the value of the environment variable {\tt CM\_LOCAL\_PATHCONFIG}.

The syntax of all configuration files is identical.  Lines are
processed from top to bottom. White space divides lines into tokens.
\begin{itemize}
\item A line with exactly two tokens associates an anchor (the first
token) with a directory in native syntax (the second token).  Neither
anchor nor directory name may contain white space and the anchor
should not contain a {\bf /}.  If the directory name is a relative
name, then it will be expanded by prepending the name of the directory
that contains the configuration file.
\item A line containing exactly one token that is the name of an
anchor cancels any existing association of that anchor with a
directory.
\item A line with a single token that consists of a single minus sign
{\bf -} cancels all existing anchors.  This typically makes sense only
at the beginning of the user-specific configuration file and
erases any settings that were made by the installation-specific
configuration file.
\item Lines with no token (i.e., empty lines) will be silently ignored.
\item Any other line is considered malformed and will cause a warning
but will otherwise be ignored.
\end{itemize}

\subsection{When to use anchor environments}

The following sample scenario for using anchor environments will
provide some background on the reasons that led to the inclusion of
this feature into make7.  In short, anchor environments will typically be
used to disambiguate between two or more different uses of the same
anchor name.  This is something that the root environment alone (i.e.,
a fixed mapping from anchors to directory names) cannot do because the
root environment binds each anchor to at most one ``meaning.''

Suppose you are developing a project {\tt P.make6} for which you obtain
two utility libraries {\tt A.make6} and {\tt B.make6} from other
programmers.  Unfortunately, since these programmers did not
coordinate their work (perhaps because they did not even know of each
other), both {\tt A.make6} and {\tt B.make6} each uses its own helper
library that is expected to be found under the name {\tt \$H/H.make6}.

Your task is to put references to {\tt A.make6} and {\tt B.make6} into {\tt
P.make6} so that everything ``works.''  Now, mentioning {\tt A.make6} and
{\tt B.make6} is easy enough, but if you do so without special
precautions, you arrive at a situation where both helpers end up being
the same---no matter how you configure the naming for anchor {\tt
\$H}.

If you have access to the description files for {\tt A.make6} and {\tt
B.make6}, you could work around this problem by changing the reference to
{\tt \$H} in one of them into something else.  Remember, however, that
in general you may not be able to do this because it could be that
either you do not have permissions to change the description files, or
you might not even have any description file that you could change
because the two libraries were given to you in ``stable'' form.

The correct solution is to add {\tt bind:}-directives where you {\em
use} {\tt A.make6} and {\tt B.make6} in your own description file {\tt P.make6}.
For example, you could write there:

\begin{verbatim}
  A.make6 (bind:(anchor:H value:$AH))
  B.make6 (bind:(anchor:H value:$BH))
\end{verbatim}

With this, the two uses of anchor {\tt \$H} will occur in different
{\em local} anchor environments.  In the example as shown, the first
such local environment effectively renames {\tt \$H} into {\tt \$AH}
while the second renames {\tt \$H} into {\tt \$BH}. Thus, you can put
independent namings for {\tt \$AH} and {\tt \$HB} into your root
configuration, these independent namings will propagate to the
respective local environment, and the two uses of {\tt \$H} will be
resolved differently---as was intended.

Another situation where anchor name clashes would definitely happen is
when two different versions (i.e., development stages) of the same set
of libraries are being used at the same time.  In this case, one
should let the libraries within each set refer to each other using
anchored names and provide (different) namings for these names using
{\tt bind:}-directives from within their respective clients.

A good way of encapsulating the construction of the required local
anchor environment for a library is to create a {\em proxy} for it
(see Section~\REF{sec:exportcalculus}).
