% -*- latex -*-

\section{Files}
\label{sec:files}

make7 uses three kinds of files to store derived information during and
between sessions:

\begin{enumerate}
\item {\it Module_Dependencies_Summary files} are used to store a highly abbreviated
version of each ML source file's abstract syntax tree---just barely
sufficient to drive make7's dependency analysis.  Module_Dependencies_Summary files are much
smaller and (for a program) easier to read than actual ML source code.
Therefore, the existence of valid module_dependencies_summary files makes make7 a lot faster
because usually most parsing operations can be avoided that way.
\item Our {\it .o7 files} 
contain executable code and a symbol table for the associated
source file.
\item {\it Library files} (sometimes called: {\em stablefiles}) contain
dependency graph, executable code, and symbol tables for an entire make7
library including all of its components (sublibraries).  Other libraries
used by a freezefile are not included in full.  Instead,
references to those libraries are recorded using their (preferably
anchored) pathnames.
\end{enumerate}

Normally, all these files are stored in a subdirectory of directory
{\tt make7}. {\tt make7} itself is a subdirectory of the directory where the
original source file or---in the case of library files---the
original make7 description file is located.

Module_Dependencies_Summary files are machine- and operating system-independent.
Therefore, they are always placed into the same directory {\tt
make7/SKEL}. Parsing (for the purpose of dependency analysis) will be
done only once even if the same file system is accessible from
machines of different type.

.o7 files and library files contain executable code and other
information that is potentially system- and architecture-dependent.
Therefore, they are stored under {\tt make7/}{\it arch}{\tt -}{\it os}
where {\it arch} is a string indicating the type of the current
CPU architecture and {\it os} a string denoting the current operating
system type.

As explained in Section~\REF{sec:stable}, library files are a bit of
an exception in the sense that they do not require any source files or
any other derived files of the same library to exist.  As a
consequence, the location of such a library file should be described
as being relative to ``the location of the original make7 description
file if that description file still existed''.  (Of course, nothing
precludes the make7 description file from actually existing, but in the
presence of a corresponding library file make7 will not take any notice
of that.)

{\em Note:} As discussed in section~\REF{sec:toolparam:cm}, make7 sometimes
looks for library files in {\tt make7/}{\it version}{\tt /}{\it arch}{\tt
-}{\it os}.  However, library files are never {\em created} there by
make7.  If several versions of the same library are to be provided, an
administrator must arrange the directory hierarchy accordingly ``by
hand''.

\subsection{Time stamps}

For module_dependencies_summary files and .o7 files, make7 uses file system time stamps
(i.e., modification time) to determine whether a file has become
outdated.  The rule is that in order to be considered ``up-to-date''
the time stamp on module_dependencies_summary file and .o7 file has to be exactly the
same\footnote{make7 explicitly sets the time stamp to be the same.} as
the one on the source file.  This guarantees that all changes to a
source will be noticed---even those that revert to an older version of
a source file.\footnote{except for the pathological case where two
different versions of the same source file have exactly the same time
stamp}

make7 also uses time stamps to decide whether tools such as Mythryl-Yacc or
Mythryl-Lex need to be run (see Section~\REF{sec:ilks}).  However, the
difference is that a file is considered outdated if it is older than
its source.  Some care on the programmers side is necessary since this
scheme does not allow make7 to detect the situation where a source file
gets replaced by an older version of itself.

\subsection{Index files}
\label{sec:indexfiles}

make7 can optionally generate a human-readable index file for each
description file.  An index files alphabetically lists all symbols
defined or imported within the given library or library component.
Index-file generation is normally disabled.  To enable it, {\tt
make7.control.generate\_index} must be set to true (see
section~\REF{sec:api}).

With index-file generation enabled, index files will be written for
all description files involved every time make7 performs a dependency
analysis.  (In other words, it is a side-effect to other make7 operations
such as {\tt make7.make} etc.)  If the name of the description files is
{\tt $p$/$d$.make6}, then the corresponding index file will be in {\tt
$p$/make7/INDEX/$d$.make6}.
