% -*- latex -*-

\section{Parallel and distributed compilation}
\label{sec:parmake}

To speed up recompilation of large projects with many source files,
makelib can exploit parallelism in the dependency graph by running
multiple compile processes in parallel.

Currently, the only kind of operating system for which this is
implemented is Unix ({\tt OPSYS\_UNIX}), where separate processes are
used.  From there, one can distribute the work across a network of
machines by taking advantage of the network file system and the
``rsh'' facility.

To perform parallel compilations, one must attach {\em compile
servers} to the currently running makelib session.  This is done using
function {\tt makelib::makelib_server::start_a_makelib_server}, which has the
following type:

\begin{verbatim}
  package makelib_server
      :
      api
          Makelib_Server;

          start_a_makelib_server
              :
              { name:       String,
                cmd:       (String, List( String ),
                path_translator: Null_Or (String -> String),
                preference:      int
              }
              ->
              Null_Or( Makelib_Server )
      end;
\end{verbatim}

Here, {\tt name} is an arbitrary string that is used by makelib when
issuing diagnostic messages concerning the server\footnote{Therefore,
it is useful to choose {\tt name} uniquely.} and {\tt cmd} is a value
suitable as argument to {\tt spawn__premicrothread::execute}.

The program to be specified by {\tt cmd} should be another instance of
makelib---running as a ``compile server''.  To start makelib as a compile
server, start {\tt mythryld} with a single command-line argument of
{\tt --compile-server}.
For example, if you have it installed in /path/to/bin/mythryld, then a
server process on the local machine could be started by

\begin{verbatim}
  makelib::makelib_server::start_a_makelib_server
      {
        name            =>  "A",
        path_translator =>  NULL,
        preference            =>  0,
        cmd             =>  ("/path/to/bin/mythryld", ["--compile-server"])
      };
\end{verbatim}

To run a process on a remote machine, e.g., ``thatmachine'', as
compute server, one can use ``rsh''.\footnote{On certain systems it
may be necessary to wrap {\tt rsh} into a script that protects rsh
from interrupt signals.}  Unfortunately, at the moment it
is necessary to specify the full path to ``rsh'' because {\tt
unix::execute} (and therefore {\tt makelib::makelib_server::start_a_makelib_server})
does not perform a {\tt PATH} search. The remote machine
must share the file system with the local machine, for example via NFS.

\begin{verbatim}
  makelib::makelib_server::start_a_makelib_server { name = "thatmachine",
                    path_translator = NULL, preference = 0,
                    cmd = ("/usr/ucb/rsh",
                           ["thatmachine",
                            "/path/to/bin/mythryld",
                            "--compile-server"]) };
\end{verbatim}

You can start as many servers as you want, but they all must have
different names.  If you attach any servers at all, then you should
attach at least two (unless you want to attach one that runs on a
machine vastly more powerful than your local one).  Local servers make
sense on multi-CPU machines: start as many servers as there are CPUs.
Parallel make is most effective on multiprocessor machines because
network latencies can have a severely limiting effect on what can be
gained in the distributed case.
(Be careful, though.  Since there is no memory-sharing to speak of
between separate instances of {\tt mythryld}, you should be sure to check
that your machine has enough main memory.)

If servers on machines of different power are attached, one can give
some preference to faster ones by setting the {\tt preference} value higher.
(But since the {\tt preference} value is consulted only in the rare case
that more than one server is idle, this will rarely lead to vastly
better throughput.) All attached servers must use the same
architecture-OS combination as the controlling machine.

In parallel mode, the master process itself normally does not compile
anything.  Therefore, if you want to utilize the master's CPU for
compilation, you should start a compile server on the same machine
that the master runs on (even if it is a uniprocessor machine).

The {\tt path_translator} argument is used when connecting to a machine with
a different file-system layout.  For local servers, it can safely be
left at {\tt NULL}.  The ``path transformation'' function is used to
translate local path names to their remote counterparts.  This can be
a bit tricky to get right, especially if the machines use automounters
or similar devices.  The {\tt path_translator} functions consumes and
produces names in makelib's internal ``protocol encoding'' (see
Section~\REF{sec:pathencode}).

Once servers have been attached, one can invoke functions like
{\tt makelib::compile}, {\tt makelib::make}, and {\tt makelib::freeze}.
They should work the way they always do, but during compilation they
will take advantage of parallelism.

When makelib is interrupted using Control-C (or such), one will sometimes
experience a certain delay if servers are currently attached and busy.
This is because the interrupt-handling code will wait for the servers
to finish what they are currently doing and bring them back to an
``idle'' state first.

\subsection{Pathname protocol encoding}
\label{sec:pathencode}

A path encoded by makelib's compile-server protocol encoding does not only
specify which file a path refers to but also, in some sense, specifies
why makelib constructed this path in the first place.  For example, the
encoding {\tt a/b/c.lib:d/e.sml} represents the file {\tt a/b/d/e.sml}
but also tells us that it was constructed by putting {\tt d/e.sml}
into the context of description file {\tt a/b/c.lib}.  Thus, an encoded
path name consists of one or more colon-separated ({\bf :}) sections,
and each section consists of slash-separated ({\bf /}) arcs.  To find
out what actual file a path refers to, it is necessary to erase all
arcs that precede colons.

The first section is special because it also specifies whether the
whole path was relative or absolute, or whether it was an anchored
path.

\begin{description}
\item[Anchored paths] start with a dollar-symbol {\bf \$}.  The name
of the anchor is the string between this leading dollar-symbol and the
first occurence of a slash {\bf /} within the first section.  The
remaining arcs of the first section are interpreted relative to the
current value of the anchor.
\item[Absolute paths] start either with a percent-sign {\bf \%} or a
slash {\bf /}.  The canonical form is the one with the percent-sign:
it specifies the volume name between the {\bf \%} and the first slash.
In the common case where the volume name is empty (i.e, {\em always} on
Unix systems), the path starts with {\bf /}.
\item[Relative paths] are all other paths.
\end{description}

Encoded path names never contain white space.  Moreover, the encoding
for path arcs, volume names, or anchor names does not contain special
characters such as {\bf /}, {\bf \$}, {\bf \%}, {\bf :}, {\bf
\verb|\|}, {\bf (}, and {\bf )}.  Instead, should white space or
special characters occur in the non-encoded name, then they will be
encoded using the escape-sequence \verb|\ddd| where {\tt ddd} is the
decimal value of the respective character's ordinal number (i.e, the
result of applying {\tt char.ord}).

The so-called {\em current} arc is encoded as {\bf .}, the {\em
parent} arc uses {\bf ..} as its representation.  It might be that
under some operating systems the names {\tt .} or {\tt ..} do not
actually refer to the current or the parent arc.  In such a case, makelib
will encode the dots in these names using the \verb|\ddd| method, too.

When issuing progress messages, makelib shows path names in a form that is
almost the same as the protocol encoding.  The only difference is that
arcs that precede colon-sign {\bf :} are enclosed within parentheses
to emphasize that they are ``not really there''.  The same form is
also used by {\tt makelib.library.descr}.

\subsection{Parallel bootstrap compilation}

The bootstrap compiler\footnote{otherwise not mentioned in this
document} with its main function {\tt make_compiler.make} and the corresponding
cross-compilation variants of the bootstrap compiler will also use any
attached compile servers.  If one intends to exclusively use the
bootstrap compiler, one can even attach servers that run on machines
with different architecture or operating system.

Since the compile-server protocol is fairly simple, it cannot handle
complicated scenarios such as the one necessary for compiling the
``init library'' (i.e., the small set of files necessary for setting up
the ``pervasive'' environment) during {\tt make_compiler.make}.  Therefore, this
will always be done locally by the master process.
