\chapter{Mythryl Source Code}

% ================================================================================
% This chapter is referenced in:
%
%     doc/tex/book.tex
%

% ================================================================================
\section{Release Notes}

\cutdef*{subsection}
\include{section-src-release-notes.tex}
\cutend*

% ================================================================================
\section{Surf Online}

If you just want to read the Mythryl source code, the easiest way is to surf 
the \ahrefloc{chapter:codebase}{\textbf{hyperlinked online HTML Mythryl codebase}}.

% ================================================================================
\section{Download Tarball}

The full Mythryl source code distribution is available from the \ahref{\mythryldownload}{\textbf{Mythryl source code download page}}.

% ================================================================================
\section{Download Debian Linux .deb Packages}

An unofficial {\tt .deb} binary distribution of the latest stable Mythryl release 
is available \ahref{\mythryldebsite}{\textbf{here}} for users of {\tt dpkg} based Linux 
distributions such as Debian and Ubuntu.

Direct download link for \textbf{Mythryl 6.1.0 .deb} package is \ahref{\mythryldebpkg}{here}.

Direct download link for \textbf{mythryl-mode.el .deb} package is \ahref{\mythrylmodedebpkg}{here}.

(Many thanks to Michele Bini for volunteering to create and maintain the Mythryl .deb distributions!)

% ================================================================================
\section{Git Repository}

If you like life on the bleeding edge you can get \ahref{\gitmythryl}{\textbf{the very latest development snapshot}} via git:

\begin{verbatim}
    git clone git://github.com/mythryl/mythryl.git
\end{verbatim}

This will give you source code only.  Since Mythryl is a self-compiling system you will also 
need a set of seed binaries to build the system.  These may be obtained initially from the 
{\tt bin/} directory of the tarball download (preceding section).

If you do not have {\tt git} installed you will have to install 
it.  Under Debian, Ubuntu and related Linux flavors this will 
involve some command like:

\begin{verbatim}
    sudo apt-get  install git-core
    sudo aptitude install git-core
    sudo synaptic install git-core
\end{verbatim}

There are a variety of git-related tools available.  On Debian Linux and its derivatives 
like Ubuntu you can try searching for them using one or more of:
\begin{verbatim}
    apt-cache search git
    apt-cache search --names-only git
    debtags search devel::rcs | grep -i git
\end{verbatim}

\begin{quotation}
\begin{quotation}
{\it \tiny Many thanks to the near-omniscient Drake Wilson for the above search hints!}
\end{quotation}
\end{quotation}


For more information on git see the \ahref{\git}{\textbf{online user manual}}.

% ================================================================================
\section{Mythryl Mode for Emacs and Xemacs}

For emacs users Phil Rand created a \textbf{mythryl-mode.el};  since his death Michele 
Bini has taken over maintainance.  Main site is \ahref{\mythrylmode}{here}; Debian 
Linux \textbf{.deb} package is \ahref{\mythrylmodedebpkg}{here}.

