\section{Line Counts}

  \begin{Table}{|l|r|r|}{align=right} \hline
                                               & Front-end & Lowcode Back-end \\ \hline
      \begin{color}{#00aa00}Generic\end{color} & 3,023 & 6,814 \\
      \begin{color}{#00aa00}Hppa\end{color}    &  725  & 2,285 \\
      \begin{color}{#00aa00}Alpha\end{color}   &  614  & 2,316 \\ \hline
     TOTAL & 4,362 & 11,415 \\ \hline
  \end{Table} 
  The table shows the number of lines involved in a basic Lowcode code
  generator for the compiler that only does graph coloring register
  allocation. The front-end column shows the number of lines specific to
  the front-end and the back-end column shows the number of lines specific to
  the Lowcode back-end. The \begin{color}{#00aa00}Generic\end{color} shows the
  number of lines that are target independent for both front-end and
  back-end. The \begin{color}{#00aa00}Hppa\end{color} and 
  \begin{color}{#00aa00}Alpha\end{color} shows the number of lines that are
  target dependent for both the HP Hppa and DEC Alpha targets.

  The bulk of the \sml{3,023} generic package to the front-end is involved in the
  generation of back-end code trees. Once this is done the incremental cost
  of adding a target is between \sml{600} to \sml{700} lines.

  The Lowcode column shows that the bulk of Lowcode is quite generic and
a client is saved from writing \sml{11,415} lines of code.

  \begin{Table}{|l|r|r|}{align=left} \hline
                & Front-End & Back-End \\ \hline
   \begin{color}{#00aa00}Generic\end{color} & 121 + 3,023 & 15,686 + 6,814\\
   \begin{color}{#00aa00}Hppa\end{color}    & 32 + 725    & 920 + 2,285 \\
   \begin{color}{#00aa00}Alpha\end{color}   & 614         & 2,316 \\ \hline
    TOTAL & 153 + 4,362 & 16,606 + 11,415 \\ \hline
  \end{Table}
  If one were to include the preliminary numbers for global acyclic
  scheduling in the above table, we find that the incremental cost
  required by the client is quite small -- approximately \sml{153}
  lines of which \sml{121} are generic. However, the scheduling
infra package is quite large, a lot of it being quite generic. 

\br{left=clear}
