\section{Line Counts}

  \begin{Table}{|l|r|r|}{align=right} \hline
                                               & frontend & Lowhalf backend \\ \hline
      \begin{color}{#00aa00}Generic\end{color} & 3,023 & 6,814 \\
      \begin{color}{#00aa00}Hppa\end{color}    &  725  & 2,285 \\
      \begin{color}{#00aa00}Alpha\end{color}   &  614  & 2,316 \\ \hline
     TOTAL & 4,362 & 11,415 \\ \hline
  \end{Table} 
  The table shows the number of lines involved in a basic Lowhalf code
  generator for the compiler that only does graph coloring register
  allocation. The frontend column shows the number of lines specific to
  the frontend and the backend column shows the number of lines specific to
  the Lowhalf backend. The \begin{color}{#00aa00}Generic\end{color} shows the
  number of lines that are target independent for both frontend and
  backend. The \begin{color}{#00aa00}Hppa\end{color} and 
  \begin{color}{#00aa00}Alpha\end{color} shows the number of lines that are
  target dependent for both the HP Hppa and DEC Alpha targets.

  The bulk of the \sml{3,023} generic package to the frontend is involved in the
  generation of backend code trees. Once this is done the incremental cost
  of adding a target is between \sml{600} to \sml{700} lines.

  The Lowhalf column shows that the bulk of Lowhalf is quite generic and
a client is saved from writing \sml{11,415} lines of code.

  \begin{Table}{|l|r|r|}{align=left} \hline
                & Frontend & Backend \\ \hline
   \begin{color}{#00aa00}Generic\end{color} & 121 + 3,023 & 15,686 + 6,814\\
   \begin{color}{#00aa00}Hppa\end{color}    & 32 + 725    & 920 + 2,285 \\
   \begin{color}{#00aa00}Alpha\end{color}   & 614         & 2,316 \\ \hline
    TOTAL & 153 + 4,362 & 16,606 + 11,415 \\ \hline
  \end{Table}
  If one were to include the preliminary numbers for global acyclic
  scheduling in the above table, we find that the incremental cost
  required by the client is quite small -- approximately \sml{153}
  lines of which \sml{121} are generic. However, the scheduling
infra package is quite large, a lot of it being quite generic. 

\br{left=clear}
